---
title: "Data Cleaning and Manipulation"
author: "Erin Rooney"
date: "2023-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r}

source("code/0-packages.R")


```

## Load Data

```{r}

catch_original = read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

#for windows typically
#catch_original = read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"))


```

## Clean Data

The select command from the dplyr package selects **columns**

The filter command filters **rows**

```{r}

#remove the All and notesRegCode columns from catch_original dataframe

catch_data =
  catch_original %>% 
  dplyr::select(-All, -notesRegCode) 

head(catch_data)
  
glimpse(catch_data)


```


Mutate Chinook so that it is numeric (not a character)

```{r}

catch_clean = 
  catch_data %>% 
  mutate(Chinook = as.numeric(Chinook))

# search for the na value

i = which(is.na(catch_clean$Chinook))

i

# located our NA value (it's an I)


catch_clean = 
  catch_data %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% 
  mutate(Chinook = as.integer(Chinook))

```


## Change shape of data use pivot longer or pivot wider

```{r}

catch_long = 
  catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), names_to = "species",
               values_to = "catch")

glimpse(catch_long)


catch_wide = 
  catch_long %>% 
  pivot_wider(names_from = "species", values_from = "catch")

glimpse(catch_wide)

```

## Practice renaming columns

```{r}
catch_long =
  catch_long %>% 
  rename(catch_thousands = catch)

glimpse(catch_long)
  
```
## Adding columns with mutate function

```{r}
catch_long =
  catch_long %>% 
  mutate(catch = catch_thousands *1000) %>% 
  dplyr::select(-catch_thousands)

glimpse(catch_long)
```

## Practice using group_by and summarise

```{r}

mean_region =
  catch_long %>% 
  group_by(Region) %>% 
  summarise(catch_mean = mean(catch))

head(mean_region)

#summarise count region

n_region =
  catch_long %>% 
  group_by(Region) %>% 
  summarise(count = n(),
            total_catch = sum(catch))

head(n_region)

```


## Lesson on filtering

```{r}

All_catch = 
  catch_long %>% 
  filter(Region %in% "ALU")

```


## Arrange some rows (use negative for largest to smallest and no negative for smallest to largest)

```{r}

mean_region =
  catch_long %>% 
  group_by(Region) %>% 
  summarise(mean_catch = mean(catch)) %>% 
  arrange(-mean_catch)

head(mean_region)

```

## Joinin' Time

load in other dataframe for left joining 

```{r}

region_defs = read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)

```

## Join our data

```{r}

catch_joined = 
  catch_long %>% 
  left_join(region_defs, by = c("Region" = "code"))

head(catch_joined)

```


```{r}

region_defs = 
  region_defs %>% 
  rename(Region = code, Region_Name = mgmtArea)

catch_joined =
  left_join(catch_long, region_defs, by = "Region")

head(catch_joined)

```

## Practice with separate() and unite() 

```{r}

sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAI-039"))

sites_df %>% 
  separate(site, c("island", "site_number"), "-")


```

```{r}

dates_df <- data.frame(year = c("1930",
                                "1930",
                                "1930"),
                       month = c("12",
                                "12",
                                "12"),
                       day = c("14",
                               "15",
                               "16"))

dates_df %>% 
  unite(date, year, month, day, sep = "-")

```


