---
title: "Geospatial Analysis with sf"
author: "Erin Rooney"
date: "2023-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load packages

```{r}
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggmap)
library(dplyr)
```

## load data

```{r}

## Read in shapefile using sf
ak_regions = read_sf("data/ak_regions_simp.shp")

plot(ak_regions)  

class(ak_regions)


```
## look at data

geometry column is a special column that contains the spacial data that is needed to map the information (multi polygon)

```{r}

head(ak_regions)

```

## CRS can be projected or unprojected
Projected = geographic coordinate system, 3D, globe
Unprojected = defines where the data is located on the earth's surface

You can view what crs is set by using the function st_crs


## Coordinate Reference Systems (CRS)

A CRS can be both unprojected and projected.

Unprojected CRS aka a geographic coordinate system (GCS)
- latitude and longitude references the earth as a 3D object
- tells us *where* something is located on the globe
- measured in angular units, i.e. degrees

Projected CRS or projected coordinated system (PCS) or projection (for short)
- latitude and longitude references the earth as a 2D object
- tells us *how* to draw or create the representation on a flat surface
- measured in linear units, i.e. meters

```{r}

st_crs(ak_regions)

```


GEOGCRS means it is unprojected and we need to transform it to add a projection to it so we can map it accurately

CRS = EPSG code that we want to use

3338 (Alaska Albers) is pretty common for mapping in Alaska

```{r}

ak_regions_3338 =
  ak_regions %>%
  st_transform(crs = 3338)

st_crs(ak_regions_3338)

```

```{r}

plot(ak_regions_3338)

```

## sf & the tidyverse

```{r}
ak_regions_3338 %>% 
  select(region)

# geometry column is a "sticky" column, it will stay with your dataframe even when not explicitly called

```
```{r}
ak_regions_3338 %>% 
  filter(region %in% "Southeast")

```

## spatial joins



```{r}

pop = read.csv("data/alaska_population.csv")


pop_4326 = st_as_sf(pop, 
                  coords = c('lng', 'lat'),
                  crs = 4326,
                  remove = F)

head(pop_4326)

```

## assign each city to a region using 'st_join()'

transform pop_4326 to crs = 3338

check the .prj file (projection) for metadata to find out crs

```{r}

pop_3338 = st_transform(pop_4326, crs = 3338)


pop_joined = 
  st_join(pop_3338, ak_regions_3338, join = st_within)

head(pop_joined)

```

## calculate total populations per region

```{r}
pop_region =
  pop_joined %>% 
  as.data.frame() %>% #turn it back into a dataframe
  group_by(region) %>% 
  summarise(total_pop = sum(population))

head(pop_region)

#dataframe and not a geometry thing because no sticky geometry column anymore

#but we do want spatial information so now we need to left_join data


#call it 3338 so we know it's a spatial object

pop_region_3338 =
  left_join(ak_regions_3338, pop_region)

plot(pop_region_3338["total_pop"])

```

Plot population by Alaska management area

```{r}

pop_mgmt_3338 = 
  pop_region_3338 %>% 
  group_by(mgmt_area) %>% 
  summarise(total_pop = sum(total_pop), do_union = FALSE)

plot(pop_mgmt_3338["total_pop"])

#write the shape file

write_sf(pop_region_3338, "output/ak_regions_population.shp", delete_layer = TRUE)


```


## visualize with ggplot

```{r}
pop_region_3338 %>% 
  ggplot()+
  geom_sf(aes(fill = total_pop))+
  labs(fill = "Total Population")+
  scale_fill_continuous(low = "#6d6875", high = "#ffcdb2", labels = comma)+
  theme_minimal()

#labels by comma!!!!!!! This is so helpful!

```

enter  ?sf::tidyverse into the console for more info


## add in rivers

```{r read rivers shape}

rivers_3338 = read_sf("data/ak_rivers_simp.shp")

st_crs(rivers_3338)

#PROJCRS not GEOGCRS means it is projected
#Albers EPSG is being used

#crs suggest r package?

```

```{r}

ggplot() +
  geom_sf(data = pop_region_3338, aes(fill = total_pop))+
  geom_sf(data = rivers_3338, aes(linewidth = StrOrder), color = "#83c5be")+
  geom_sf(data = pop_3338, aes(), size = .5)+
  scale_linewidth(range = c(0.01, 0.2), guide = F)+
  labs(fill = "Total Population")+
  scale_fill_continuous(low = "#6d6875", high = "#ffcdb2", labels = comma)+
  theme_minimal()

```

## Incorporate base maps into static maps ussing 'ggmap'

```{r}

source("code/6-function_mapboundingbox.R")


pop_3857 =
  pop_3338 %>% 
  st_transform(crs = 3857)


```

set up base map and bounding box

```{r}
bbox = c(-170, 52, -130, 64) #roughly southern alaska

ak_map = get_stamenmap(bbox, zoom = 4)

ak_map_3857 = ggmap_bbox_to_3857(ak_map)
```

plot map

```{r}
ggmap(ak_map_3857)+
  geom_sf(data = pop_3857, aes(color = population), inherit.aes = F)+
    scale_color_continuous(low = "#6d6875", high = "#ffcdb2", labels = comma)

```

## Interactive maps with Leaflet

```{r}

epsg3338 = leaflet::leafletCRS(
  crsClass = "L.Proj.CRS",
  code = "EPSG:3338",
  proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
  resolutions = 2^(16:7))

```

transform pop region 

```{r}

pop_region_4326 = 
  pop_region_3338 %>% 
  st_transform(crs = 4326)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
        addPolygons(data = pop_region_4326, 
                    fillColor = "gray",
                    weight = 1)

m

```

