---
title: "RMarkdown_practice2"
author: "Erin Rooney"
date: "2023-02-28"
output: html_document
---

# North Pole Environmental Bottle Chemistry Introduction

[@universityoftexasatelpaso2018]

[Here is the link to the data!](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X "Click here")

Seawater chemistry and complementary conductivity, temperature, and depth data taken from the Arctic Ocean (near the North Pole).

### Insert Packages

```{r message=FALSE, results='hide'}
#knit('myfile.rmd', tangle=TRUE)
source('code/0-packages.R')
```


### Read Data

```{r}


BGchem = read.csv("data/BGchem2008data.csv")


```

### Summarise Data

```{r}
colnames(BGchem)

str(BGchem)

head(BGchem)

summary(BGchem)

view(BGchem)
```


### Analysis

### Calculate summary statistics
```{r stats, echo = FALSE}

nitrate <- mean(BGchem$NO3)
nitrite <- mean(BGchem$NO2)
amm <- mean(BGchem$NH4)
phos <- mean(BGchem$P)

```



### Calculate mean Redfield ratio
```{r}

ratio <- (nitrate + nitrite + amm)/phos

```

The Redfield ratio in this data set is approximately 'r round(ratio)'

### Plot Redfield ratio

```{r}

plot(BGchem$P, BGchem$NO2 + BGchem$NO3 + BGchem$P)

```

```{r BGchem, echo = FALSE}

ggplot(BGchem, aes(x = CTD_Temperature, y = CTD_Depth, color = P))+
  geom_point()+
  scale_y_reverse()+
  theme_bw()

```
